---
title: \vspace{4cm} **\Large Intensification Mechanisms Driving Dietary Change among the Great Plains Big
  Game Hunters of North America**\vspace{1cm}
date: "Last compiled on `r format(Sys.time(), '%A %B %d %Y, %X, %Z')`"
author:
  - \*Erik Otárola-Castillo$^1$
  - Melissa G. Torquato$^1$
  - \*Matthew E. Hill$^2$
indent: yes
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage[labelfont=bf,width=.8\textwidth]{caption}
    - \captionsetup[figure]{labelfont=bf}
    - \usepackage[pagewise]{lineno}
output: 
  pdf_document: 
    latex_engine: xelatex
  word_document: default
  html_document:
    df_print: paged
link-citations: yes
bibliography: IntensificationGreatPlainsRMarkdown.bib
biblio-style: apalike
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(list(scipen=5, digits = 2))
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
########################################
# check for required packages          #
########################################
tinytex::install_tinytex()
list.of.packages <- c("betareg", "MASS", "car", "emmeans", "brms", "dplyr", "tidyr", "stringr","ggplot2","knitr", "kableExtra", "tinytex")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
########################################
# load packages                        #
########################################
require(kableExtra)
require(tinytex)
require(betareg)
require(MASS)
require(car)
require(emmeans)
require(brms)
require(dplyr)
require(tidyr)
require(stringr)
require(ggplot2)
require(knitr)

########################################
# load LATEX files into temp dir       #
########################################
tempsty <- file.path(tempdir(), "unicode-math.sty")
      file.copy("unicode-math.sty", tempsty, overwrite = TRUE)
      tempxesty <- file.path(tempdir(), "unicode-math-xetex.sty")
      file.copy("unicode-math-xetex.sty", tempxesty, overwrite = TRUE)
      tempxesty1 <- file.path(tempdir(), "unicode-math-table.tex")
      file.copy("unicode-math-table.tex", tempxesty, overwrite = TRUE)
      tempxesty2 <- file.path(tempdir(), "unicode-math-luatex.sty")
      file.copy("unicode-math-luatex.sty", tempxesty, overwrite = TRUE)

```

\noindent
$^1$Department of Anthropology, Purdue University, West Lafayette, IN, USA, email:\href{mailto:eoc@purdue.edu}{\color{blue} \,eoc@purdue.edu}\newline
$^2$Department of Anthropology, University of Iowa, Iowa City, IA, USA, email:\href{mailto:matthew-ehill@uiowa.edu}{\color{blue} \,matthew-e-hill@uiowa.edu}\newline

\vspace{2.5cm}
\textbf{*\*Corresponding Authors}

\vspace{4cm}
\noindent
Manuscript in press: Defining and measuring diversity in archaeology. Forthcoming volume edited by M. Eren and B. Buchanan. Berghahn.  

\linenumbers
\newpage
```{r child= "rmd_files/introduction.Rmd"}

```
\newpage
```{r child= "rmd_files/section-1.Rmd"}

```
\newpage
```{r child= "rmd_files/section-2.Rmd"}

```
\newpage
```{r child= "rmd_files/section-3.Rmd"}

```
\newpage
```{r child= "rmd_files/results.Rmd"}

```
\newpage
```{r child= "rmd_files/discussion.Rmd"}

```
\newpage
```{r child= "rmd_files/conclusion.Rmd"}

```
\newpage
```{r child= "rmd_files/acknowledgments.Rmd"}

```
\newpage

## \hfil REFERENCES CITED\hfil

\parindent 0pt

\parskip 1.5ex

\onehalfspacing

<div id="refs"></div>

\newpage

```{r child= "rmd_files/tables.Rmd"}

```

